---
title: "KNN_Exosolar_Analysis"
author: "Niket"
output: rmarkdown::github_document
---

KNN Model - When we need classification for a new data, the KNN algorithm goes through the entire dataset to find the k-nearest values to the new data values, or the k number of values most similar to the new record, and then outputs the mode (most frequent class) for a classification problem. The value of k is specified by user.

```{r}
# load the new dataset
exosim <- read.csv(file.choose(), header = TRUE) # load exos_new-imp2.csv
```

```{r}
#### Converting variable Probability_of_life to factor
exosim$Probability_of_life <- as.factor(exosim$Probability_of_life)
table(exosim$Probability_of_life)
# So we have 328 planets in our dataset with some probability of life.
```

#### Create a normalization function
```{r}
# we can normalize the data 

normalize <- function(x) {
return((x - min(x))/(max(x) - min(x)))} # create normalize function
```

#### KNN

```{r}
exosim_numerical <- subset(exosim, select = c(PlanetaryMassJpt, RadiusJpt, PeriodDays, SemiMajorAxisAU, Eccentricity, PeriastronDeg, LongitudeDeg, AscendingNodeDeg, InclinationDeg, SurfaceTempK, AgeGyr, DiscoveryYear, DistFromSunParsec, HostStarMassSlrMass, HostStarRadiusSlrRad, HostStarMetallicity, HostStarTempK, HostStarAgeGyr))
```

```{r}
#install.packages("corrr")
library(corrr)
exosim_numerical %>% correlate() %>% rplot()
```

We see very less correlation between the variables, so we don't have to remove any.

```{r}
#### Converting variable Probability_of_life to factor
exosim$Probability_of_life <- as.factor(exosim$Probability_of_life)
table(exosim$Probability_of_life)
# So we have 328 planets in our dataset with some probability of life.
```


```{r}
label_1 <- exosim$Probability_of_life # our variable for classification
```


```{r}
# Categorical variables need to be represented with numbers 
TypeFlag_dummies <- model.matrix( ~ TypeFlag - 1, data = exosim)

DiscoveryMethod_dummies <- model.matrix( ~ DiscoveryMethod - 1, data = exosim)

ListsPlanetIsOn_dummies <- model.matrix( ~ ListsPlanetIsOn - 1, data = exosim)
```

```{r}
exosim_numcat <- data.frame(exosim_numerical, TypeFlag_dummies, DiscoveryMethod_dummies, ListsPlanetIsOn_dummies)
```

```{r}
str(exosim_numcat)
```

```{r}
# normalize 
norm_exosim_numcat <- as.data.frame(sapply(exosim_numcat, normalize))
```

```{r}
set.seed(1234)
oneortwo <- sample(1:2 , length(exosim$PlanetIdentifier), replace = TRUE, prob=c(0.8, 0.2)) # generating random values and storing them
```

```{r}
# create train data frame
train_1 <- norm_exosim_numcat[oneortwo == 1, ]

# create test data frame
test_1 <- norm_exosim_numcat[oneortwo == 2, ]

# create data frame to apply train and test upon
train_1_label <- label_1[oneortwo == 1]
test_1_label <- label_1[oneortwo == 2]
```

```{r}
require(class)  # to use knn algorithm  

# splitting the data
set.seed(1234)
life_predicted_1 <- knn(train = train_1, test = test_1, cl = train_1_label, k = 20)
```


```{r}
results_1 <- data.frame(life_predicted_1, test_1_label)
#install.packages('caret')
require(caret)
```

```{r}
#install.packages('e1071')
require(e1071)
```

```{r}
accuracy_1 <- paste("Accuracy of KNN Model is:", sum(life_predicted_1 == test_1_label)/length(life_predicted_1))
knn <- sum(life_predicted_1 == test_1_label)/length(life_predicted_1)

confusionMatrix(table(results_1))
```

We get an accuracy of 91.48% which is really good, but has a room for improvemnet.

***